Studying public transit API query logs\\ to get an indication of travel flows

### 2016-01-07 paper.tex

%Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
%This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
% Context & % Need
In the field of urban planning, researchers need an indication of how people move between cities. 
Yet, getting statistics of travel flows within public transit systems has proven to be troublesome.
% Task
In order to get an indication of travel flows between cities in Belgium,
we analyzed the query logs of the iRail API, a highly expressive route planning API for the Belgian railways.
We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.56\%$ and $1.66\%$ of the amount of monthly passengers.
% Object & Findings
We illustrate that we can see morning and evening peaks in the query logs.
Moreover, using origin-destination visualizations, we can confirm that Brussels, the capital, acts as a central hub.
The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
The findings correspond to the real travel demand, according to experts of the passenger federation Trein Tram Bus.
% Conclusion
We conclude that query logs of route planners are of high importance in getting an indication of travel flows.
% Perspectives
However, better analytics would be acquirable using dedicated HTTP POST requests to capture travel intentions.
%%Redundant according to RV:
%Highly expressive transport data publishing methods such as route planning API exist, as well as data dumps or data fragments.
%In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.

@@

%IRail is a not for profit organization that stimulates digital creativity concerning mobility.
This interface is open source and we were able to suggest changes to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
As of the 17th of December 2015, iRail publishes the most recent 1,000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
The JSON objects can be read one per one, so UNIX commands such as \emph{grep} work well with this format.
We chose this over CSV as we have more flexibility towards changing the object properties in the future.
Next, we chose to add an HTTP header with a link to the query log context according to the JSON-LD specification\footnote{\url{}} to turn it into Linked Data.
%This is the JSON-LD context of all the objects in the, JSON-stream. Subsequentially, we call this a \emph{JSONLD-stream} and this has been formalized at \emph{https://github.com/pietercolpaert/jsonld-stream}.

@@

%\todo{Maybe a sub header here?} - PC: I wouldn't do a subheader...
We are particularly interested in studying the potential application of \emph{route planning} queries as an information resource for flow analysis due to the input it captures from human users.
The other features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
We made a copy of this dataset in CSV format available at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
%CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
Each row within this dataset contains:
\begin{enumerate}
\item A timestamp of when the request was executed,
\item The request path between two stations with longitudes and latitudes,
\item The user-agent string cfr. RFC2616,
\item The operating system of the user-agent and
\item Whether or not the user-agent is a bot.
\end{enumerate}

@@

%After analysis, it appeared 10\% of the request where performed by Google bots.
%\todo{add glue to Method section?}

@@

Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
Several stages of data processing are required to arrive at this outcome.
While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may over represent connections between certain stations.
In this regard, such queries have been excluded to improve the accuracy of our analysis. 
Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
Spelling mistakes, and the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) was employed to reduce spelling variations and link this with a geographical location.

@@

There are several inconsistent patterns in the data that do not correspond with an existing understanding of flows in Belgium.
In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API that were not flagged as bots or a high concentration people who utilise route planning applications on a regular basis.
Accordingly, the reverse is identified in Liège where an unusually large connection originating from a suburb towards Liege city.
This prompted speculation, as well as discussions, on how such outliers should be treated or what may have led to their existence.

>>>>

%Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
%This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
% Context & % Need
In the field of urban planning, researchers need an indication of how people move between cities. 
Yet, getting statistics of travel flows within public transit systems has proven to be troublesome.
% Task
In order to get an indication of travel flows between cities in Belgium,
we analyzed the query logs of the iRail API, a highly expressive route planning API for the Belgian railways.
We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.56\%$ and $1.66\%$ of the amount of monthly passengers.
% Object & Findings
Using data visualizations, we illustrate the commuting patterns in Belgium and confirm that Brussels, the capital, acts as a central hub. 
The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
The findings correspond to the real travel demand, according to experts of the passenger federation Trein Tram Bus.
% Conclusion
We conclude that query logs of route planners are of high importance in getting an indication of travel flows.
% Perspectives
However, better analytics would be acquirable using dedicated HTTP POST requests to capture travel intentions.
%%Redundant according to RV:
%Highly expressive transport data publishing methods such as route planning API exist, as well as data dumps or data fragments.
%In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.

@@

%IRail is a not for profit organization that stimulates digital creativity concerning mobility.
This interface is open source and we were able to suggest changes to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
As of the 17th of December 2015, iRail publishes the most recent 1,000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
The JSON objects can be read one per one, so UNIX commands such as \emph{grep} work well with this format.
We chose this over CSV as we have more flexibility towards changing the object properties in the future.
Next, we chose to add an HTTP header with a link to the query log context according to the JSON-LD specification\footnote{\url{http://www.w3.org/TR/json-ld/}} to turn it into Linked Data.
%This is the JSON-LD context of all the objects in the, JSON-stream. Subsequentially, we call this a \emph{JSONLD-stream} and this has been formalized at \emph{https://github.com/pietercolpaert/jsonld-stream}.

@@

%\todo{Maybe a sub header here?} - PC: I wouldn't do a subheader...
We are particularly interested in studying the potential application of \emph{route planning} queries as an information resource for flow analysis due to the input it captures from human users.
The other features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
We made a copy of this dataset available in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
%CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
Each row within this dataset contains:
\begin{enumerate}
\item A timestamp of when the request was executed,
\item The request path between two stations with longitudes and latitudes,
\item The user-agent string cfr. RFC2616,
\item The operating system of the user-agent and
\item Whether or not the user-agent is a bot.
\end{enumerate}

@@

%After analysis, it appeared 10\% of the request where performed by Google bots.

@@

Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
Several stages of data processing are required to arrive at this outcome.
%As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may over represent connections between certain stations.
In this regard, such queries have been excluded to improve the accuracy of our analysis. 
Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
Spelling mistakes, and the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) was employed to reduce spelling variations and link this with a geographical location.

@@

There are several inconsistent patterns in the data that do not correspond with an existing understanding of flows in Belgium.
In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API that were not flagged as bots or a high concentration people who utilise route planning applications on a regular basis.
Alternatively the large number of trips may have originated from users who repeatedly request for updates on delayed trains or from a concentration of people who utilise route planning applications on a regular basis.
Accordingly, the reverse is identified in Liège where an unusually large connection originating from a suburb towards Liege city.
This prompted speculation, as well as discussions, on how such outliers should be treated or what may have led to their existence.
As with any form of analysis based on appropriation of data, we acknowledge that more work is required to identify caveats and understand the validity of the findings presented.
